---
title: "Methylation Microarray Chip Randomization"
author: "Samuel Prates dos Santos"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: darkly
    highlight: zenburn
    df_print: paged
    code_folding: show
---

# Introduction

This R Markdown document performs a **stratified randomization** of biological
samples across methylation microarray chips.  It is designed to work with
**any Excel spreadsheet** — just update the [Configuration](#configuration)
section to match your file.

The algorithm enforces four guarantees:

1. **Paired tissues** (e.g., a tumour *CPT* and its matched adjacent normal
   *CPN* from the same patient) are placed in the **same chip column**.
2. **Vertical constraint**: within each column the tissue type of the sample
   at row 1 and the tissue type of the sample at the last row are
   **different**.
3. **Horizontal constraint**: for every row, no two horizontally adjacent
   samples (neighbouring columns) share the **same tissue type**.
4. **Stratified balance**: every variable you list in `stratify_cols` is
   distributed proportionally across all chips using
   [OSAT](https://bioconductor.org/packages/OSAT/) (Optimal Sample Assignment
   Tool), which optimises **all variables simultaneously** via a score-based
   greedy search.

---

# Configuration {#configuration}

> ✏️ **Edit only this chunk** to adapt the script to your spreadsheet.

```{r configuration, message = FALSE}
library(here)

# ── 1. File path ──────────────────────────────────────────────────────────────
# ** IMPORTANT: replace "your_file.xlsx" with your actual file name. **
# Build the path relative to the project root with here().
# Switch to read.csv(file_path) in the data-loading chunk if using a CSV file.
file_path <- here("Spreadsheets", "your_file.xlsx")

# ── 2. Chip dimensions ────────────────────────────────────────────────────────
chip_rows <- 8L   # rows per chip   (e.g. 8 for the Illumina EPIC array)
chip_cols <- 6L   # columns per chip (e.g. 6 for the Illumina EPIC array)

# ── 3. Tissue-type column ─────────────────────────────────────────────────────
# Name of the column that stores the tissue-type label (e.g. "CPT" or "CPN").
# Used to enforce the vertical and horizontal constraints.
# Set to NULL if your data has no tissue-type distinction.
tissue_col <- "Tissue_Type"

# ── 4. Pairing column ─────────────────────────────────────────────────────────
# Name of the column that links a tumour to its matched normal (e.g. a patient
# ID).  All samples that share the same value are placed in the SAME column.
# Set to NULL if you have no paired samples.
pair_col <- "Paired_ID"

# ── 5. Stratification variables ───────────────────────────────────────────────
# Vector of column names whose categories you want balanced across chips.
# Example: c("Staging", "Sex") keeps the same proportion of every staging
# category and every sex value on each chip.
stratify_cols <- c("Staging", "Age", "Sex")

# ── 6. Reproducibility ────────────────────────────────────────────────────────
# Change this number to obtain a different valid randomization.
seed <- 123L
```

---

# Libraries

```{r libraries, message = FALSE, warning = FALSE}
library(readxl)    # read Excel files
library(tidyverse) # dplyr, tidyr, stringr, purrr, ggplot2 …
library(ggplot2)   # plots
library(gridExtra) # arrange multiple plots side by side
library(scales)    # percent labels on axes
library(here)      # portable file paths
# OSAT — install once from Bioconductor if needed:
#   if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
#   BiocManager::install("OSAT")
if (!requireNamespace("OSAT", quietly = TRUE))
  stop(
    "Package 'OSAT' is not installed.\n",
    "Install it from Bioconductor with:\n",
    "  BiocManager::install(\"OSAT\")"
  )
library(OSAT)
```

---

# Data Loading

```{r data_loading}
# read_excel() loads the first sheet of the workbook.
# Replace with read.csv(file_path) if your data is a CSV file.
if (!file.exists(file_path)) {
  stop(
    "File not found: ", file_path,
    "\n  -> Open the Configuration chunk and set file_path to your spreadsheet."
  )
}
data_raw <- read_excel(file_path)

cat("Dimensions  :", nrow(data_raw), "rows x", ncol(data_raw), "columns\n")
cat("Column names:", paste(names(data_raw), collapse = ", "), "\n\n")
head(data_raw)
```

---

# Column Validation

```{r column_validation}
# Check that every configured column actually exists in the loaded spreadsheet.
# All issues are collected first so the error message lists everything at once.
issues <- character(0L)

if (!is.null(tissue_col) && !(tissue_col %in% names(data_raw)))
  issues <- c(issues,
    paste0("  tissue_col    = \"", tissue_col, "\"  <- NOT FOUND"))

if (!is.null(pair_col) && !(pair_col %in% names(data_raw)))
  issues <- c(issues,
    paste0("  pair_col      = \"", pair_col, "\"  <- NOT FOUND"))

bad_strat <- setdiff(stratify_cols, names(data_raw))
if (length(bad_strat) > 0L)
  issues <- c(issues,
    paste0("  stratify_cols contains: ",
           paste(paste0('"', bad_strat, '"'), collapse = ", "),
           "  <- NOT FOUND"))

if (length(issues) > 0L) {
  stop(
    "\nColumn validation FAILED — configured columns not found in the spreadsheet:\n",
    paste(issues, collapse = "\n"),
    "\n\nColumns available in the spreadsheet:\n  ",
    paste(names(data_raw), collapse = ", "),
    "\n\nOpen the Configuration chunk and correct the column names."
  )
}

cat("Column validation passed — all configured columns found.\n")
```

---

# Data Exploration

```{r data_exploration}
# ── Tissue-type distribution ──────────────────────────────────────────────────
if (!is.null(tissue_col) && tissue_col %in% names(data_raw)) {
  cat("=== Distribution of '", tissue_col, "' ===\n", sep = "")
  print(table(data_raw[[tissue_col]], useNA = "ifany"))
  cat("\n")
}

# ── Stratification-variable distributions ─────────────────────────────────────
for (col in stratify_cols) {
  if (col %in% names(data_raw)) {
    cat("=== Distribution of '", col, "' ===\n", sep = "")
    print(table(data_raw[[col]], useNA = "ifany"))
    cat("\n")
  } else {
    warning("Column '", col, "' not found. Check your stratify_cols setting.")
  }
}

# ── Chip-capacity summary ─────────────────────────────────────────────────────
spots_per_chip <- chip_rows * chip_cols
n_samples      <- nrow(data_raw)
n_chips        <- ceiling(n_samples / spots_per_chip)
n_empty        <- n_chips * spots_per_chip - n_samples

cat("=== Chip Capacity Summary ===\n")
cat("Samples                 :", n_samples,      "\n")
cat("Spots per chip          :", spots_per_chip, "\n")
cat("Chips needed            :", n_chips,        "\n")
if (n_empty > 0L) cat("Empty spots (last chip) :", n_empty, "\n")
```

---

# Randomization

## Helper Functions

```{r helper_functions}
# ─────────────────────────────────────────────────────────────────────────────
# order_column_samples()
#
# Shuffles a set of sample IDs so that the tissue type at position 1 differs
# from the tissue type at the last position — the VERTICAL constraint.
#
# Arguments:
#   sample_ids : integer vector of row IDs for this column
#   tissues    : character vector of tissue types (same length as sample_ids)
#   max_tries  : maximum number of random shuffles before the fallback
#
# Returns: sample_ids reordered to satisfy (or approximate) the constraint.
# ─────────────────────────────────────────────────────────────────────────────
order_column_samples <- function(sample_ids, tissues, max_tries = 2000L) {
  n <- length(sample_ids)
  if (n <= 1L) return(sample_ids)

  # If every tissue is identical the constraint is impossible; just shuffle.
  if (length(unique(na.omit(tissues))) <= 1L) return(sample(sample_ids))

  for (attempt in seq_len(max_tries)) {
    perm    <- sample(n)
    t_first <- tissues[perm[1L]]
    t_last  <- tissues[perm[n]]
    if (is.na(t_first) || is.na(t_last) || t_first != t_last) {
      return(sample_ids[perm])
    }
  }

  # Guaranteed fallback: swap the last position with the first element that
  # has a different tissue type from position 1.
  perm     <- sample(n)
  t_first  <- tissues[perm[1L]]
  diff_pos <- which(tissues[perm] != t_first & !is.na(tissues[perm]))
  if (length(diff_pos) > 0L) {
    swap           <- diff_pos[length(diff_pos)]
    perm[c(swap, n)] <- perm[c(n, swap)]
  }

  warning("Vertical constraint could not be satisfied after ", max_tries,
          " tries. Returning best approximation.")
  sample_ids[perm]
}

# ─────────────────────────────────────────────────────────────────────────────
# count_h_violations()
#
# Counts pairs of horizontally adjacent cells (same row, neighbouring columns)
# that share the same non-NA tissue type — violations of the HORIZONTAL
# constraint.
#
# Arguments:
#   tissue_matrix : character matrix [chip_rows x chip_cols]
#
# Returns: integer count of violations.
# ─────────────────────────────────────────────────────────────────────────────
count_h_violations <- function(tissue_matrix) {
  if (ncol(tissue_matrix) < 2L) return(0L)
  count <- 0L
  for (r in seq_len(nrow(tissue_matrix))) {
    for (c in seq_len(ncol(tissue_matrix) - 1L)) {
      t1 <- tissue_matrix[r, c]
      t2 <- tissue_matrix[r, c + 1L]
      if (!is.na(t1) && !is.na(t2) && t1 == t2) count <- count + 1L
    }
  }
  count
}
```

## Main Algorithm

```{r randomization}
set.seed(seed)

# ── Step 1: Add an internal row identifier ────────────────────────────────────
data <- data_raw %>% mutate(.sid = row_number())

# ── Step 2: Separate paired from unpaired samples ────────────────────────────
if (!is.null(pair_col) && pair_col %in% names(data)) {

  pair_counts     <- data %>%
    filter(!is.na(.data[[pair_col]])) %>%
    count(.data[[pair_col]], name = "n")

  paired_patients <- pair_counts %>%
    filter(n > 1L) %>%
    pull(1)

  paired_data   <- data %>% filter(.data[[pair_col]] %in% paired_patients)
  unpaired_data <- data %>%
    filter(!(.data[[pair_col]] %in% paired_patients) |
             is.na(.data[[pair_col]]))

  cat("Paired samples   :", nrow(paired_data),
      "(from", length(paired_patients), "patients)\n")
  cat("Unpaired samples :", nrow(unpaired_data), "\n")

} else {
  paired_data     <- data[0L, ]
  unpaired_data   <- data
  paired_patients <- character(0L)
  cat("No pairing column — all", nrow(unpaired_data),
      "samples treated as unpaired.\n")
}

# ── Step 3: Initialise chip-column slots and place paired samples ─────────────
# Paired samples are placed first so each patient's CPT + CPN share a column.
# col_assign[[chip_i]][[col_j]] holds the .sid values for that position.
col_assign <- lapply(seq_len(n_chips), function(i) {
  lapply(seq_len(chip_cols), function(j) integer(0L))
})

chip_i <- 1L
col_j  <- 1L

advance_pointer <- function() {
  col_j  <<- col_j + 1L
  if (col_j > chip_cols) {
    col_j  <<- 1L
    chip_i <<- chip_i + 1L
  }
}

if (length(paired_patients) > 0L) {
  for (patient in sample(paired_patients)) {
    group_sids <- paired_data %>%
      filter(.data[[pair_col]] == patient) %>%
      pull(.sid)
    n_group <- length(group_sids)

    # Skip columns that cannot fit the entire pair group
    while (chip_i <= n_chips &&
           length(col_assign[[chip_i]][[col_j]]) + n_group > chip_rows) {
      advance_pointer()
    }
    if (chip_i > n_chips) {
      warning("Not enough chip space for all paired groups!")
      break
    }
    col_assign[[chip_i]][[col_j]] <-
      c(col_assign[[chip_i]][[col_j]], group_sids)
  }
}

# ── Step 4: OSAT — assign unpaired samples to chips ──────────────────────────
# OSAT (Optimal Sample Assignment Tool, Bioconductor) balances ALL variables in
# stratify_cols simultaneously using a score-based greedy optimisation.
# It replaces the earlier single-variable round-robin approach.
#
# How it works:
#   a. setup.sample()    — declares which columns to balance.
#   b. setup.container() — describes the chip layout (spots_per_chip per chip,
#                          n_chips chips total).
#   c. OSAT()            — finds an assignment that minimises imbalance.
#   d. getSample()       — returns the original rows + their chip assignment.
if (nrow(unpaired_data) > 0L && length(stratify_cols) > 0L) {

  # OSAT requires a plain data.frame; stratification columns must be factors
  osat_input <- unpaired_data %>%
    select(all_of(stratify_cols)) %>%
    mutate(across(everything(), as.factor)) %>%
    as.data.frame()

  osat_s        <- setup.sample(osat_input, optimal = stratify_cols)

  # One-chip template: spots_per_chip positions (OSAT replicates it n_chips times)
  osat_template <- data.frame(Position = seq_len(spots_per_chip))
  osat_nc       <- setup.container(osat_template, n = n_chips, batch = "Chip")

  osat_res <- OSAT(osat_s, osat_nc, method = "greedy")

  # getSample() returns one row per sample (original input order) with the
  # assigned Chip number added as a column.
  osat_assigned <- getSample(osat_res)

  # Sort unpaired samples so that chip 1's samples fill the queue first,
  # chip 2 second, etc. — the filling loop in Step 5 uses this ordering.
  unpaired_sorted <- unpaired_data[order(osat_assigned$Chip), ]

} else {
  unpaired_sorted <- unpaired_data[sample(nrow(unpaired_data)), ]
}

# ── Step 5: Fill remaining chip-column slots with OSAT-ordered samples ────────
unpaired_queue <- unpaired_sorted$.sid
queue_idx      <- 1L

for (ci in seq_len(n_chips)) {
  for (cj in seq_len(chip_cols)) {
    free <- chip_rows - length(col_assign[[ci]][[cj]])
    while (free > 0L && queue_idx <= length(unpaired_queue)) {
      col_assign[[ci]][[cj]] <-
        c(col_assign[[ci]][[cj]], unpaired_queue[queue_idx])
      queue_idx <- queue_idx + 1L
      free      <- free - 1L
    }
  }
}

# ── Step 6: Order samples within each column — VERTICAL constraint ────────────
# Within every column the tissue type at row 1 must differ from the tissue
# type at the last occupied row.
for (ci in seq_len(n_chips)) {
  for (cj in seq_len(chip_cols)) {
    sids <- col_assign[[ci]][[cj]]
    if (length(sids) == 0L) next

    if (!is.null(tissue_col) && tissue_col %in% names(data)) {
      tissues                <- data[[tissue_col]][match(sids, data$.sid)]
      col_assign[[ci]][[cj]] <- order_column_samples(sids, tissues)
    } else {
      col_assign[[ci]][[cj]] <- sample(sids)
    }
  }
}

# ── Step 7: Optimise column order per chip — HORIZONTAL constraint ─────────────
# Try 300 random column permutations per chip; keep the arrangement with the
# fewest same-tissue-type neighbours in the same row.
if (!is.null(tissue_col) && tissue_col %in% names(data)) {
  for (ci in seq_len(n_chips)) {

    build_tissue_matrix <- function(perm) {
      mat <- matrix(NA_character_, nrow = chip_rows, ncol = chip_cols)
      for (j in seq_len(chip_cols)) {
        sids <- col_assign[[ci]][[perm[j]]]
        if (length(sids) > 0L) {
          ts <- data[[tissue_col]][match(sids, data$.sid)]
          mat[seq_along(ts), j] <- ts
        }
      }
      mat
    }

    best_perm  <- seq_len(chip_cols)
    best_score <- count_h_violations(build_tissue_matrix(best_perm))

    for (trial in seq_len(300L)) {
      perm  <- sample(chip_cols)
      score <- count_h_violations(build_tissue_matrix(perm))
      if (score < best_score) {
        best_score <- score
        best_perm  <- perm
      }
      if (best_score == 0L) break
    }

    col_assign[[ci]] <- col_assign[[ci]][best_perm]
  }
}

# ── Step 8: Flatten everything into a tidy result data frame ──────────────────
result_rows <- vector("list", n_chips * chip_cols * chip_rows)
row_count   <- 0L

for (ci in seq_len(n_chips)) {
  for (cj in seq_len(chip_cols)) {
    sids <- col_assign[[ci]][[cj]]
    for (rk in seq_along(sids)) {
      row_count <- row_count + 1L
      result_rows[[row_count]] <- tibble(
        .sid     = sids[rk],
        chip_num = ci,
        chip_row = as.integer(rk),
        chip_col = as.integer(cj)
      )
    }
  }
}

chip_layout <- bind_rows(result_rows[seq_len(row_count)])

# Join position information back to the full sample data
randomized_data <- chip_layout %>%
  left_join(data, by = ".sid") %>%
  select(-.sid) %>%
  arrange(chip_num, chip_col, chip_row)

cat("Randomization complete! Total samples placed:", nrow(randomized_data), "\n")
```

## Constraint Validation

```{r validation}
if (!is.null(tissue_col) && tissue_col %in% names(randomized_data)) {

  # ── Vertical constraint ────────────────────────────────────────────────────
  vertical_check <- randomized_data %>%
    group_by(chip_num, chip_col) %>%
    arrange(chip_row, .by_group = TRUE) %>%
    summarise(
      first_tissue = first(.data[[tissue_col]]),
      last_tissue  = last(.data[[tissue_col]]),
      .groups = "drop"
    ) %>%
    mutate(
      violation = !is.na(first_tissue) & !is.na(last_tissue) &
                   first_tissue == last_tissue
    )

  n_vert <- sum(vertical_check$violation)
  if (n_vert == 0L) {
    cat("Vertical constraint   : No violations\n")
  } else {
    cat("Vertical constraint   :", n_vert, "violation(s)\n")
    print(vertical_check %>% filter(violation))
  }

  # ── Horizontal constraint ──────────────────────────────────────────────────
  horiz_check <- randomized_data %>%
    group_by(chip_num, chip_row) %>%
    arrange(chip_col, .by_group = TRUE) %>%
    summarise(tissues = list(.data[[tissue_col]]), .groups = "drop") %>%
    rowwise() %>%
    mutate(
      n_violations = {
        ts <- unlist(tissues)
        if (length(ts) < 2L) 0L
        else sum(ts[-length(ts)] == ts[-1L] &
                 !is.na(ts[-length(ts)]) & !is.na(ts[-1L]),
                 na.rm = TRUE)
      }
    ) %>%
    ungroup()

  n_horiz <- sum(horiz_check$n_violations)
  if (n_horiz == 0L) {
    cat("Horizontal constraint : No violations\n")
  } else {
    cat("Horizontal constraint :", n_horiz, "adjacent violation(s)\n")
    print(horiz_check %>% filter(n_violations > 0L) %>% select(-tissues))
  }
}

# ── Pairing constraint ─────────────────────────────────────────────────────────
if (!is.null(pair_col) && pair_col %in% names(randomized_data) &&
    length(paired_patients) > 0L) {

  pair_check <- randomized_data %>%
    filter(.data[[pair_col]] %in% paired_patients) %>%
    group_by(chip_num, .data[[pair_col]]) %>%
    summarise(n_cols = n_distinct(chip_col), .groups = "drop") %>%
    filter(n_cols > 1L)

  if (nrow(pair_check) == 0L) {
    cat("Pairing constraint    : All paired samples in the same column\n")
  } else {
    cat("Pairing constraint    :", nrow(pair_check), "violation(s)\n")
    print(pair_check)
  }
}
```

---

# Visualisation

## Chip Layout Grid

```{r plot_chip_layout, fig.width = 12, fig.height = 5}
# Colour-code each cell by tissue type (or by the first stratification variable
# if no tissue_col is set).
color_col <- if (!is.null(tissue_col) && tissue_col %in% names(randomized_data)) {
  tissue_col
} else if (length(stratify_cols) > 0L &&
           stratify_cols[1L] %in% names(randomized_data)) {
  stratify_cols[1L]
} else {
  NULL
}

if (!is.null(color_col)) {
  chip_plots <- lapply(seq_len(n_chips), function(ci) {
    chip_data <- randomized_data %>%
      filter(chip_num == ci) %>%
      mutate(
        chip_row = factor(chip_row, levels = rev(seq_len(chip_rows))),
        chip_col = factor(chip_col, levels = seq_len(chip_cols))
      )

    ggplot(chip_data,
           aes(x = chip_col, y = chip_row, fill = .data[[color_col]])) +
      geom_tile(colour = "white", linewidth = 1.2) +
      scale_fill_brewer(palette = "Set2", na.value = "grey70",
                        name = color_col) +
      labs(title = paste("Chip", ci), x = "Column", y = "Row") +
      theme_minimal(base_size = 11) +
      theme(
        plot.title      = element_text(hjust = 0.5, face = "bold"),
        legend.position = "bottom"
      )
  })

  do.call(grid.arrange, c(chip_plots, ncol = min(3L, n_chips)))
}
```

## Proportion of Each Category per Chip

```{r plot_proportion_per_chip, fig.width = 9, fig.height = 4}
# Build the list of variables to visualise: tissue type + all stratification
# variables (duplicates removed, NULL values ignored).
cols_to_plot <- unique(c(
  if (!is.null(tissue_col)) tissue_col,
  stratify_cols
))

# One stacked bar chart per variable.
# X-axis = chip number; Y-axis = proportion of each category on that chip.
for (col in cols_to_plot) {
  if (!(col %in% names(randomized_data))) next

  prop_chip <- randomized_data %>%
    filter(!is.na(.data[[col]])) %>%
    count(chip_num, .data[[col]]) %>%
    group_by(chip_num) %>%
    mutate(proportion = n / sum(n)) %>%
    ungroup()

  p <- ggplot(prop_chip,
              aes(x = factor(chip_num), y = proportion,
                  fill = factor(.data[[col]]))) +
    geom_bar(stat = "identity", position = "fill") +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    scale_fill_brewer(palette = "Set2", name = col) +
    labs(
      title = paste("Proportion of", col, "per Chip"),
      x     = "Chip",
      y     = "Proportion"
    ) +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))

  print(p)
}
```

## Proportion of Each Category per Row Position

```{r plot_proportion_per_row, fig.width = 9, fig.height = 4}
# One stacked bar chart per variable.
# X-axis = row position (1 … chip_rows); Y-axis = proportion aggregated
# across all chips — shows whether any row position is biased toward a
# particular category.
for (col in cols_to_plot) {
  if (!(col %in% names(randomized_data))) next

  prop_row <- randomized_data %>%
    filter(!is.na(.data[[col]])) %>%
    count(chip_row, .data[[col]]) %>%
    group_by(chip_row) %>%
    mutate(proportion = n / sum(n)) %>%
    ungroup()

  p <- ggplot(prop_row,
              aes(x = factor(chip_row), y = proportion,
                  fill = factor(.data[[col]]))) +
    geom_bar(stat = "identity", position = "fill") +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    scale_fill_brewer(palette = "Set2", name = col) +
    labs(
      title = paste("Proportion of", col, "per Row Position (all chips)"),
      x     = "Row",
      y     = "Proportion"
    ) +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))

  print(p)
}
```

## Proportion of Each Category per Column Position

```{r plot_proportion_per_col, fig.width = 9, fig.height = 4}
# One stacked bar chart per variable.
# X-axis = column position (1 … chip_cols); Y-axis = proportion aggregated
# across all chips — shows whether any column position is biased toward a
# particular category.
for (col in cols_to_plot) {
  if (!(col %in% names(randomized_data))) next

  prop_col <- randomized_data %>%
    filter(!is.na(.data[[col]])) %>%
    count(chip_col, .data[[col]]) %>%
    group_by(chip_col) %>%
    mutate(proportion = n / sum(n)) %>%
    ungroup()

  p <- ggplot(prop_col,
              aes(x = factor(chip_col), y = proportion,
                  fill = factor(.data[[col]]))) +
    geom_bar(stat = "identity", position = "fill") +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    scale_fill_brewer(palette = "Set2", name = col) +
    labs(
      title = paste("Proportion of", col, "per Column Position (all chips)"),
      x     = "Column",
      y     = "Proportion"
    ) +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))

  print(p)
}
```

---

# Results

```{r results}
cat("=== RANDOMIZATION SUMMARY ===\n")
cat("Total samples randomized        :", nrow(randomized_data), "\n")
cat("Number of chips used            :", n_chips,        "\n")
cat("Spots per chip                  :", spots_per_chip, "\n")
if (n_empty > 0L) cat("Empty spots on last chip        :", n_empty, "\n")

# Full chip-assignment table (paginated in HTML output)
randomized_data %>%
  select(chip_num, chip_col, chip_row, everything()) %>%
  arrange(chip_num, chip_col, chip_row)
```

---

# Export

```{r export}
# Save the chip-assignment table as a CSV file in the Spreadsheets/ folder.
# The file name includes today's date so repeated runs do not overwrite each
# other.
output_path <- here(
  "Spreadsheets",
  paste0("Randomized_Chips_", format(Sys.time(), "%Y-%m-%d_%H-%M-%S"), ".csv")
)

write.csv(randomized_data, output_path, row.names = FALSE)
cat("Results saved to:", output_path, "\n")
```
